---
title: "BM1"
output: html_document
editor_options: 
  chunk_output_type: console
---

The following code analysis the breast metastatic sample with metastasis to the liver and to the pleural effusion.
This works as an example on how to employ CopyKit for the analysis of multiple datasets.

# BM1

## Reading datasets
``` {r bm1_data, class.source="bg-success"}
# Running data with CopyKit for the primary breast sample
bm1_breast <-
  runVarbin(
    "/mnt/lab/users/dminussi/projects/CopyKit_Manuscript_Code/datasets/BM1/breast/marked_bams/",
    remove_Y = TRUE
  )

# Finding diploid and low-quality cells and excluding it from the copykit object
bm1_breast <- findOutliers(bm1_breast, resolution = 0.8)
bm1_breast <- findAneuploidCells(bm1_breast)

bm1_breast <- bm1_breast[, colData(bm1_breast)$outlier == FALSE]
bm1_breast <- bm1_breast[, colData(bm1_breast)$is_aneuploid == TRUE]

# Adding the tissue information to colData
colData(bm1_breast)$timepoint <- 'breast'

# ~~~~~~~~~~~~~~~~~~~~~~~

# Running data with CopyKit for the liver metastasis sample
bm1_liver <-
  runVarbin(
    "/mnt/lab/users/dminussi/projects/CopyKit_Manuscript_Code/datasets/BM1/liver/marked_bams/",
    remove_Y = TRUE
  )

# Finding diploid and low-quality cells and excluding it from the copykit object
bm1_liver <- findOutliers(bm1_liver)
bm1_liver <- findAneuploidCells(bm1_liver)

bm1_liver <- bm1_liver[, colData(bm1_liver)$outlier == FALSE]
bm1_liver <- bm1_liver[, colData(bm1_liver)$is_aneuploid == TRUE]

# Adding the tissue information to colData
colData(bm1_liver)$timepoint <- 'liver'

# Running data with CopyKit for the pleural effusion metastasis sample
bm1_pleural <-
  runVarbin(
    "/mnt/lab/users/dminussi/projects/CopyKit_Manuscript_Code/datasets/BM1/pleural/marked_bams/",
    remove_Y = TRUE
  )

# Finding diploid and low-quality cells and excluding it from the copykit object
bm1_pleural <- findOutliers(bm1_pleural)
bm1_pleural <- findAneuploidCells(bm1_pleural)

bm1_pleural <- bm1_pleural[, colData(bm1_pleural)$outlier == FALSE]
bm1_pleural <-
  bm1_pleural[, colData(bm1_pleural)$is_aneuploid == TRUE]

# Adding the tissue information to colData
colData(bm1_pleural)$timepoint <- 'pleural'
```

Merging the three datasets
``` {r bm1_merging, class.source="bg-success"}
# Merging the three copykit objects
bm1_merged <- cbind(bm1_breast,
                    bm1_liver,
                    bm1_pleural)

```

From here on the analysis follow the same steps as a standard CopyKit workflow analysis.

## Running UMAP and Clustering
``` {r bm1_umap_cl, class.source="bg-success"}
bm1_merged <- runUmap(bm1_merged)

bm1_merged <- findSuggestedK(bm1_merged)
bm1_merged_suggestedk <- plotSuggestedK(bm1_merged)
bm1_merged_suggestedk

bm1_merged <- findClusters(bm1_merged)

bm1_merged <- calcConsensus(bm1_merged)
bm1_merged <- runConsensusPhylo(bm1_merged)

plotHeatmap(bm1_merged, label = c('subclones', 'timepoint'))

bm1_merged_umap_p <- plotUmap(bm1_merged, label = 'subclones')
bm1_merged_umap_p

# Cluster c3 from the pleural sample is a cluster of tumor-normal doublets
# We can subset out of the CopyKit object in a similar way to the filtering
# functions with the information from colData
bm1_merged <- bm1_merged[, colData(bm1_merged)$subclones != 'c3']

# Re-clustering the sample after doublet removal
# This sample has a smaller sample size, therefore we are reducing the
# n_neighbors parameter from the UMAP and increasing min_dist
bm1_merged <- runUmap(bm1_merged, n_neighbors = 10, min_dist = 0.1)

# Grid Search of Jaccard Similarity (cluster stability)
bm1_merged <- findSuggestedK(bm1_merged)
bm1_merged <- findClusters(bm1_merged)

# Plotting the UMAP colored by the tissue of origin from the colData information
bm1_merged_tp_umap_p <- plotUmap(bm1_merged, label = 'timepoint')
bm1_merged_tp_umap_p

# Plotting the UMAP colored by the subclones from the colData information
bm1_merged_umap_p <- plotUmap(bm1_merged, label = 'subclones')
bm1_merged_umap_p
```

## Consensus tree

To root the tree, we will use an inferred Most Recent Common Ancestral from the primary tumor and provide that as an argument to the runConsensusPhylo function.
This consensus tree will be used by plotHeatmap to order the subclones on the plot
``` {r bm1_tree, class.source="bg-success"}
# The primary sample per se has very few cells so we will add a subclone 
# information to the colData to use the later inferMrca function in CopyKit
colData(bm1_breast)$subclones <- 'c1'
bm1_breast <- calcConsensus(bm1_breast)

# Inferring the MRCA on the primary breast sample
bm1_breast <- inferMrca(bm1_breast)

# calculating the consensus of the Merged dataset and using the inferred 
# primary MRCA as the root of the tree
bm1_merged <- calcConsensus(bm1_merged)
bm1_merged <- runConsensusPhylo(bm1_merged,
                                root = 'user',
                                root_user = metadata(bm1_breast)$inferred_mrca)
# Rotating branches
consensusPhylo(bm1_merged) <-
  ape::rotate(consensusPhylo(bm1_merged), 6)

bm1_merged_consensus_phylo <-
  plotPhylo(bm1_merged, label = 'subclones', consensus = TRUE)
bm1_merged_consensus_phylo

# Calculating cophenetic distances between subclones
ape::cophenetic.phylo(consensusPhylo(bm1_merged))
```

``` {r bm1_heatmap, fig.width = 8, fig.height = 9, class.source="bg-success"}
# Plotting the copy number heatmap with annotation data from the subclones and
# the tissue of origin
plotHeatmap(bm1_merged, label = c('subclones', 'timepoint'))
```

``` {r bm1_consensus_heatmap_genes, fig.width = 9, class.source="bg-success"}
# Plotting a consensus heatmap with the plotHeatmap function. 
# The annotation represents the subclones and relevant cancer genes are marked.
plotHeatmap(
  bm1_merged,
  label = 'subclones',
  consensus = TRUE,
  genes = c(
    "MYC",
    "MYB",
    "BRCA1",
    "ERBB2",
    "CDH1",
    "FGFR1",
    "AKT2",
    "CDK4",
    "CCNE1",
    "CCND1",
    "MTOR",
    "FGF10",
    "BRAF",
    "AURKA"
  )
)

# Using the plotGeneCopy function to plot the segment ratios mean values for 
# the selected genes and coloring it by the tissue of origin from the metadata
bm1_merged_gc <-
  plotGeneCopy(
    bm1_merged,
    genes = c(
      "MYC",
      "MYB",
      "BRCA1",
      "ERBB2",
      "CDH1",
      "FGFR1",
      "AKT2",
      "CDK4",
      "CCNE1",
      "CCND1",
      "MTOR",
      "FGF10",
      "BRAF",
      "AURKA"
    ),
    label = 'timepoint',
    dodge.width = .8
  ) +
  scale_fill_hue()
bm1_merged_gc
```



