---
title: ""
output: html_document
editor_options:
  chunk_output_type: console
---

# CM1 and CM2

## CM1

## CM1 colon
```{r cm1_primary, class.source="bg-success"}
# Running data with CopyKit for the primary colon sample
cm1_tumor_primary <- runVarbin("/mnt/lab/users/dminussi/projects/CopyKit_Manuscript_Code/datasets/CM1/colon/marked_bams", remove_Y = TRUE)

# Finding diploid and low-quality cells and excluding it from the copykit object
cm1_tumor_primary <- findOutliers(cm1_tumor_primary)
cm1_tumor_primary <- findAneuploidCells(cm1_tumor_primary)

cm1_tumor_primary <- cm1_tumor_primary[,colData(cm1_tumor_primary)$outlier == FALSE]
cm1_tumor_primary <- cm1_tumor_primary[,colData(cm1_tumor_primary)$is_aneuploid == TRUE]

# Adding the tissue information to colData
colData(cm1_tumor_primary)$timepoint <- 'primary'
```

## CM1 liver
``` {r cm1_met, class.source="bg-success"}
# Running data with CopyKit for the liver sample
cm1_tumor_met <-
  runVarbin(
    "/mnt/lab/users/dminussi/projects/CopyKit_Manuscript_Code/datasets/CM1/liver/marked_bams",
    remove_Y = TRUE
  )

# Finding diploid and low-quality cells and excluding it from the copykit object
cm1_tumor_met <- findOutliers(cm1_tumor_met)
cm1_tumor_met <- findAneuploidCells(cm1_tumor_met)

cm1_tumor_met <-
  cm1_tumor_met[, colData(cm1_tumor_met)$outlier == FALSE]
cm1_tumor_met <-
  cm1_tumor_met[, colData(cm1_tumor_met)$is_aneuploid == TRUE]

# Adding the tissue information to colData
colData(cm1_tumor_met)$timepoint <- 'metastasis'
```

## CM1 merging
``` {r cm1_merge, class.source="bg-success"}
# Merging the three copykit objects
cm1_merged <- cbind(cm1_tumor_primary,
                    cm1_tumor_met)

cm1_merged <- runUmap(cm1_merged)

cm1_merged <- findSuggestedK(cm1_merged)
cm1_merged_suggestedk <- plotSuggestedK(cm1_merged)
cm1_merged_suggestedk

cm1_merged <- findClusters(cm1_merged)

# HDBSCAN is an outlier aware clustering algorithm
# in this analysis all cells marked as outliers (c0) from hdbscan are excluded.
cm1_merged <- cm1_merged[, colData(cm1_merged)$subclones != 'c0']

cm1_merged_tp_umap_p <- plotUmap(cm1_merged, label = 'timepoint')
cm1_merged_tp_umap_p

cm1_merged_umap_p <- plotUmap(cm1_merged, label = 'subclones')
cm1_merged_umap_p

cm1_merged <- calcConsensus(cm1_merged)
cm1_merged <- runConsensusPhylo(cm1_merged)

colData(cm1_merged)$timepoint <-
  forcats::fct_relevel(colData(cm1_merged)$timepoint,
                       c("primary", "metastasis"))

cm1_selected_genes = c(
  "SMAD3",
  "FHIT",
  "APC",
  "SOX4",
  "IGFBP7",
  "CDK8",
  "PIK3CA",
  "MYC",
  'TP53',
  "GATA4",
  "CHEK1",
  "TGFB1",
  "TIAM1"
)

cm1_merged_selected_hvg_gc <-
  plotGeneCopy(cm1_merged,
               genes = cm1_selected_genes,
               label = 'timepoint',
               dodge.width = .8) + scale_fill_hue(direction = -1)
cm1_merged_selected_hvg_gc
```

To root the tree, we will use an inferred Most Recent Common Ancestral from the primary tumor and provide that as an argument to the runConsensusPhylo function.
This consensus tree will be used by plotHeatmap to order the subclones on the plot
``` {r cm1_tree, class.source="bg-success"}
# clustering and inferring the MRCA from the clusters
cm1_tumor_primary <- runUmap(cm1_tumor_primary)
cm1_tumor_primary <- findSuggestedK(cm1_tumor_primary)
cm1_tumor_primary <- findClusters(cm1_tumor_primary)

# HDBSCAN is an outlier aware clustering algorithm
# in this analysis all cells marked as outliers (c0) from hdbscan are excluded.
cm1_tumor_primary <-
  cm1_tumor_primary[, colData(cm1_tumor_primary)$subclones != 'c0']


# calculating the consensus of the Merged dataset and using the inferred
# primary MRCA as the root of the tree
cm1_tumor_primary <- calcConsensus(cm1_tumor_primary)
cm1_tumor_primary <- inferMrca(cm1_tumor_primary)
cm1_merged <- runConsensusPhylo(
  cm1_merged,
  root = 'user',
  root_user = metadata(cm1_tumor_primary)$inferred_mrca
)

# relevel factors to plot in the desired order
colData(cm1_merged)$timepoint <-
  forcats::fct_relevel(colData(cm1_merged)$timepoint,
                       c("metastasis", "primary"))

# plotting the phylogeny with subclones labels and pie charts
#indicating the frequency of each timepoint
cm1_merged_consensus_phylo <-
  plotPhylo(cm1_merged,
            label = 'subclones',
            consensus = TRUE,
            group = 'timepoint')
cm1_merged_consensus_phylo
```

``` {r cm1_heatmap, fig.width = 8, fig.height = 9, class.source="bg-success"}
plotHeatmap(cm1_merged, label = c('subclones', 'timepoint'))
```

```{r cm1_consensus_heatmap, class.source="bg-success"}
plotHeatmap(
  cm1_merged,
  label = c('subclones'),
  consensus = TRUE,
  genes = cm1_selected_genes
)
```


## CM2

## CM2 colon
```{r cm2_primary, class.source="bg-success"}
# Running data with CopyKit for the primary colon sample
cm2_tumor_primary <- runVarbin("/mnt/lab/users/dminussi/projects/CopyKit_Manuscript_Code/datasets/CM2/colon/marked_bams", remove_Y = TRUE)

# Finding diploid and low-quality cells and excluding it from the copykit object
cm2_tumor_primary <- findOutliers(cm2_tumor_primary)
cm2_tumor_primary <- findAneuploidCells(cm2_tumor_primary)

cm2_tumor_primary <- cm2_tumor_primary[,colData(cm2_tumor_primary)$outlier == FALSE]
cm2_tumor_primary <- cm2_tumor_primary[,colData(cm2_tumor_primary)$is_aneuploid == TRUE]

# Adding the tissue information to colData
colData(cm2_tumor_primary)$timepoint <- 'primary'
```

## CM2 liver
``` {r cm2_met, class.source="bg-success"}
# Running data with CopyKit for the liver sample
cm2_tumor_met <-
  runVarbin(
    "/mnt/lab/users/dminussi/projects/CopyKit_Manuscript_Code/datasets/CM2/liver/marked_bams",
    remove_Y = TRUE
  )

# Finding diploid and low-quality cells and excluding it from the copykit object
cm2_tumor_met <- findOutliers(cm2_tumor_met)
cm2_tumor_met <- findAneuploidCells(cm2_tumor_met)

cm2_tumor_met <-
  cm2_tumor_met[, colData(cm2_tumor_met)$outlier == FALSE]
cm2_tumor_met <-
  cm2_tumor_met[, colData(cm2_tumor_met)$is_aneuploid == TRUE]

# Adding the tissue information to colData
colData(cm2_tumor_met)$timepoint <- 'metastasis'
```

## CM2 merging
``` {r cm2_merge, class.source="bg-success"}
# Merging the three copykit objects
cm2_merged <- cbind(cm2_tumor_primary,
                    cm2_tumor_met)

cm2_merged <- runUmap(cm2_merged)

cm2_merged <- findSuggestedK(cm2_merged)
cm2_merged_suggestedk <- plotSuggestedK(cm2_merged)
cm2_merged_suggestedk

cm2_merged <- findClusters(cm2_merged)

# HDBSCAN is an outlier aware clustering algorithm
# in this analysis all cells marked as outliers (c0) from hdbscan are excluded.
cm2_merged <- cm2_merged[, colData(cm2_merged)$subclones != 'c0']

cm2_merged_tp_umap_p <- plotUmap(cm2_merged, label = 'timepoint')
cm2_merged_tp_umap_p

cm2_merged_umap_p <- plotUmap(cm2_merged, label = 'subclones')
cm2_merged_umap_p

cm2_merged <- calcConsensus(cm2_merged)
cm2_merged <- runConsensusPhylo(cm2_merged)

colData(cm2_merged)$timepoint <-
  forcats::fct_relevel(colData(cm2_merged)$timepoint,
                       c("primary", "metastasis"))

cm2_selected_genes = c(
  "IGF1R",
  "SMAD6",
  "FGFR2",
  "STAT3",
  "SOX9",
  "KLF5",
  "FGF9",
  "MSH2",
  "APC",
  "TP53",
  "TNFRSF6B",
  "CHEK2",
  "MAP3K8"
)

cm2_merged_selected_hvg_gc <-
  plotGeneCopy(cm2_merged,
               genes = cm2_selected_genes,
               label = 'timepoint',
               dodge.width = .8) + scale_fill_hue(direction = -1)
cm2_merged_selected_hvg_gc
```

To root the tree, we will use an inferred Most Recent Common Ancestral from the primary tumor and provide that as an argument to the runConsensusPhylo function.
This consensus tree will be used by plotHeatmap to order the subclones on the plot
``` {r cm2_tree, class.source="bg-success"}
# clustering and inferring the MRCA from the clusters
cm2_tumor_primary <- runUmap(cm2_tumor_primary)
cm2_tumor_primary <- findSuggestedK(cm2_tumor_primary)
cm2_tumor_primary <- findClusters(cm2_tumor_primary)

# HDBSCAN is an outlier aware clustering algorithm
# in this analysis all cells marked as outliers (c0) from hdbscan are excluded.
cm2_tumor_primary <-
  cm2_tumor_primary[, colData(cm2_tumor_primary)$subclones != 'c0']

# calculating the consensus of the Merged dataset and using the inferred
# primary MRCA as the root of the tree
cm2_tumor_primary <- calcConsensus(cm2_tumor_primary)
cm2_tumor_primary <- inferMrca(cm2_tumor_primary, value = 0.8)
cm2_merged <- runConsensusPhylo(
  cm2_merged,
  root = 'user',
  root_user = metadata(cm2_tumor_primary)$inferred_mrca
)
consensusPhylo(cm2_merged) <- phytools::rotateNodes(consensusPhylo(cm2_merged), c(25, 27))

# relevel factors to plot in the desired order
colData(cm2_merged)$timepoint <-
  forcats::fct_relevel(colData(cm2_merged)$timepoint,
                       c("metastasis", "primary"))

# plotting the phylogeny with subclones labels and pie charts
#indicating the frequency of each timepoint
cm2_merged_consensus_phylo <-
  plotPhylo(cm2_merged,
            label = 'subclones',
            consensus = TRUE,
            group = 'timepoint')
cm2_merged_consensus_phylo
```

``` {r cm2_heatmap, fig.width = 8, fig.height = 9, class.source="bg-success"}
plotHeatmap(cm2_merged, label = c('subclones', 'timepoint'))
```

```{r cm2_consensus_heatmap, class.source="bg-success"}
plotHeatmap(
  cm2_merged,
  label = c('subclones'),
  consensus = TRUE,
  genes = cm2_selected_genes
)
```

